Non-word repetition in Yélî Dnye

### 2020-12-23 manuscript.Rmd
pass in progress, knits

abstract: >
   In nonword repetition (NWR) studies, participants are presented auditorily with an item that is phonologically legal but lexically meaningless in their language. NWR scores are thought to reflect long-term phonological knowledge as well as online phonological working memory and flexible production patterns. In this study, we report on NWR results among children learning Yélî Dnye, an isolate spoken on Rossel Island in Papua New Guinea that has an unusually dense phonological inventory. This study contributes to three lines of research. First, we document that non-word items containing typologically frequent sounds are repeated without changes more often that non-words containing typologically rare sounds. Second, we document rather weak effects of item length, contributing to other research suggesting that length effects may be language-specific. Third, we do not find strong individual variation effects in this population, contrary to previous results documenting strong age-related effects. Together, these data provide a unique view of online phonological processing in an understudied language while making preliminary connections between patterns in language development and in the frequency of linguistic features in human language at large.

@@

Children's perception and production of phonetic and phonological units continues developing well beyond the first year of life, even extending into middle childhood [e.g., @hazan2000development]. Much of the evidence for later phonological development comes through the use of nonword repetition (NWR) tasks. In a NWR task, participants hear a short word-like form that is phonologically legal but lexically meaningless in the language(s) they are learning. After hearing this non-word, the participant's task is to try to immediately and precisely repeat it. NWR scores are thought to reflect long-term phonological knowledge (to perceive the item precisely despite not having heard it before) as well as online phonological working memory (to encode the item in the interval between hearing it and saying it back) and flexible production patterns (to produce the item precisely despite not having pronounced it before).

NWR has been used to seek answers to a variety of theoretical questions, including what the links are between phonology, working memory, and the lexicon [@bowey2001nonword], and how extensively phonological constraints found in the lexicon affect online production [@gallagher2014identity]. NWR is also frequently used in applied contexts, notably as a diagnostic tool for language delays and disorders [@estes2007differences]. Since non-words can be generated in any language, it has attracted the attention of researchers working in multilingual and linguistically diverse environments, particularly in Europe [@meir2016disentangling;@is08042009language]. However, it has seldom been used outside of Europe and North America [with exceptions including @gallagher2014identity and @cristia2020infant]. Further, the languages it has been used with have historically had small to moderate phonological inventories [@maddieson2005correlating].

<!--M2A: Is the motivating idea here that application to YD is of interest because it's very different from what we've look at with NWR in the past? Or is there a specific hypothesis you had in mind about how inventory size affects phonological development and, hence, NWR performance? I could come up with something, but I think you had an idea already and I don't want to step on it!-->

In the present study, we use NWR to investigate the phonological development of children learning Yélî Dnye, an isolate language spoken in Papua New Guinea (PNG) that has a large and unusually dense phonological inventory. The study was designed to contribute to multiple aspects of our understanding of phonological development: First, we included a subset of non-word items with typologically rare and/or challenging sounds to ask whether these rare sounds are disadvantaged in perception and/or production, both in terms of overall repetition scores and patterns of mispronunciation. Second, we varied the length (in syllables) of non-words to investigate the impact of word length on repetition and mispronunciation; previous NWR research has shown variable effects of length between populations. Third, these data give a first impression of how this task is responded to in a rarely-studied language and cultural setting; we add to this contribution by further investigating whether there are structured sources that account for individual variation in our sample.
<!--M2A: This is also a note to myself; it seems we could use some more references in the above!-->

Before jumping into the details of our study design we first give an overview of Yélî Dnye phonology as well as a brief ethnographic review of the developmental environment on Rossel Island. Since NWR has been almost exclusively used in urban, industrialized populations [see @cristia2020infant for an exception], we provide this additional ethographic information to contextualize the adaptations we have made in running the task and in gathering age and other demographic information, compared to what is typical in urban, industrialized settings.

@@

In total, Yélî Dnye uses 90 distinctive segments (not including an additional 3 rarely used consonants), far outstripping the phonemic inventory size of other documented Papuan languages [@foley1986papuan; @levinsonYDgrammar] and includes at least two contrasts not yet been documented elsewhere [labial-coronal double-articulations with dental vs. post-alveolar coronal placement for both oral and nasal stops, @ladefoged1996sounds; @maddiesonIPphoneticsYD].

Most words in Yélî Dnye are bisyllabic (~50%), with monosyllabic words (~40%) appearing most commonly after that, and with tri-and-above syllabic words appearing least frequently [~10%; based on >5800 lexemes in the most recent Yélî Dnye dictionary at time of writing; @levinsonYDgrammar]. The vast majority of syllables use a CV format. A small portion of the lexicon features words with a final closed syllable, limited to codas of -/m/, -/p/, or -/j/ (e.g., 'ndap' (Spondylus shell) /ṇṭæp/). However, in spontaneous speech an epenthetic /ɯ/ is often inserted after word-final coda, yielding a CV.CV structure instead (e.g., 'ndapî' /'ṇṭæ.pɯ/). This process is used frequently with English loan words that have a coda (e.g., 'ponî' (phone) /'pɔ.nɯ/). An even smaller portion of the lexicon features words starting with a vowel (e.g., 'ala' (here) /æ.'læ/), but these are limited to /æ/-. Finally, the lexicon features a handful of single-vowel grammatical morphemes (see @levinsonYDgrammar for details).

@@

Most speakers of Yélî Dnye grow up speaking it monolingually until they begin attending school around the age of 7; school instruction is in English. While monolingual Yélî Dnye upbringing is common, multilingual families are not unusual, particularly in the region around the Catholic Mission---the same region in which the current data were collected---where there is a higher incidence of married-in mothers from other islands [@brownIPchildrearing]. Children in these multilingual families grow up speaking Yélî Dnye plus English, Tok Pisin, and/or another language(s) from the region.

@@

Laying 250 nautical miles off the coast of mainland PNG and surrounded by a barrier reef, transport to and from the island is both infrequent and irregular. There is no electricity system (we use solar panels) or motor vehicles on the island, save dinghys with outboard motors that are primarily reserved for medical transport when there is diesel available. There is a medical radio connection at the health clinic, located at the Catholic Mission, via which news and messages are often passed to and from mainland PNG. There is also a cellular tower on the island, though it is often inoperative, usually for multiple months or years at a time. As such, access to news and outside connections, including international phone calls and digital exchanges that require significant data transfer, is typically unavailable. Our data collection is therefore limited to the duration of the researchers' on-island visits.

Despite restricted outside contact, formal education is a priority for many Yélî families. A recent study surveying more than 40 parents of young children in this region of the island found that nearly all parents completed 6 or more years of education, with 65% of mothers and 49% of fathers completing the full range of locally available education (~9 years), with around half of those parents having pursued an additional 2+ years of education on other islands in the region or on mainland PNG [@casillas2020early]. While there are a handful of local schools around the island, these may be well out of walking distance for many children (i.e., more than 1 hour on foot or by canoe each day). It is very common for households that are situated close to a school to billet their school-aged relatives during the weekdays (sometimes the weekends too, if the child's home is very far) for long segements of each school year. Combined with practices of collective child guardianship within each close-knit hamlet, the practical consequence of this billeting for researchers is that adult consent can often come from a combination of aunts, uncles, adult cousins, and grandparents standing in for the child's biological parents. On top of this, child assent is culturally pertinent, as child independence is encouraged and respected from toddlerhood onward [@brownIPchildrearing].^[For these reasons, we were unable to record precise details for a few children, including child date of birth, parent age in years, and parent education.] Finally, as mentioned, the language of schooling in this region is English, and so the majority of teens and adults on the island speak some English---many speak it very fluently.

While work on Yélî language development is growing [e.g., @brown2011cultural; @brown2014interactional; @brownIPchildrearing; @casillas2020early; @liszkowski2012prelinguistic], our knowledge of children's linguistic development is quite limited, and research into their phonological development in particular has only just begun [e.g., @peuteIPconsonants]. What we do know about the early language environment is that, while direct speech to children under 3;0 is relatively infrequent throughout the day [@casillas2020early], shared caregiving practices and a near-universal fondness for social interaction with young children ensure that speech directed to children comes from all types of speakers: women, men, and other children [@bunceURcrosscultural; @casillas2020early]. While speech from women predominates in young children's speech environments on Rossel, as it does elsewhere [@bergelsoncasillas2019what; @bunceURcrosscultural], there is a significant and marked increase in child-directed speech from other children as infants get older [@bunceURcrosscultural; @casillas2020early]. This increase in child-directed speech from other children is attributed to the fact that, starting around age 2, children often spend large swaths of the day playing, swimming, and foraging for fruit, nuts, and shellfish in large (~10 members) independent and mixed-age child play groups [@brownIPchildrearing; @casillas2020early].

@@

In a basic NWR task, the participant listens to a production of a word-like form, such as /bilik/, and then repeats back what they heard without changing any phonological feature that is contrastive in the language. For instance, in English, a response of [bilig] or [pilik] would be scored as incorrect; a response [biːlik], where the vowel is lengthened without change of quality would be scored as correct, because English does not have a general feature of contrastive vowel length. There is some variation in how past NWR studies have designed the presentation procedure and structure of items. For example, while items are often presented orally by the experimenter [@torrington2015non], an increasing number of studies have turned instead to playing back pre-recorded stimuli in order to increase control in stimulus presentation [@brandeker2015language]. Additionally, while some studies have used 10-15 non-words, others have employed up to 46 unique items [@piazzalunga2019articulatory]. Authors also often modulate structural complexity, typically measured in terms of item length (measured in number of syllables) and/or syllable structure [open as opposed to closed syllables, @gallon2007non].

Previous work typically steers clear of articulatorily and/or acoustically challenging sounds, but we included some in our experiment to more adequately represent Yélî Dnye's phonology. We ultimately used a relatively large number of items that would enable us to explore both variation in structural complexity and in more vs. less challenging sounds. However, aware that this large item inventory might render the task longer and more tiresome, we split items across children (see below). Naturally, designing the task in this way may make the study of individual variation within the population more difficult because different children are exposed to different items. However, a review of previous work on individual variation suggested to us that effects of individual differences in NWR are relatively small, and would not be detectable with the sample size that we could collect during our short visit. That said, we contribute to the literature by also reporting descriptive analyses of individual variation that could potentially be integrated in meta-analytic efforts.

Based on previous work, we looked at potential changes with age [@farmani2018normalization; @kalnak2014nonword; @vance2005speech]<!--, and differences relating to bilingual language exposure [@brandeker2015language; @meir2016disentangling; @meir2017independent]-->. Previous work typically finds no significant differences as a function of maternal education [e.g., @farmani2018normalization; @kalnak2014nonword; @meir2017independent] or child gender [@chiat2007preschool]. Although past research has not often investigated potential effects of birth order on NWR, there is a sizable literature on these effects in other language tasks [@havron2019effect], and therefore we report on these too.

@@

- Does the cross-linguistic frequency of sounds in the stimuli predict NWR scores? Are rarer sounds more often substituted by commoner sounds?
- How does score change as a function of item length in number of syllables?
- Is individual variation in NWR scores attributable to child age, sex, birth order, and/or maternal education?

In view of the hypothesis-driven nature of this work, we had considered boosting the interpretational value of this evidence by announcing our analysis plans prior to conducting them. However, we realized that even pre-registering an analysis would be equivocal because we would not have enough power to look at all relationships of interest, in many cases possibly not enough to detect any of the known effects, given their variability across studies. To illustrate this issue, we portray in Figure 1 studies in which children's NWR scores were gathered between 4 and 12 years of age, and reported separately for items that are relatively short (1-2 syllables) versus longer items (3-4 syllables). Notice that the effect of stimulus length is minuscule among Italian children [@piazzalunga2019articulatory], but considerable among Tsimane' children  [@cristia2020infant], where a drop of 40 percentage points is observed at all ages. A similar large difference in NWR scores for short versus long items was observed among Arabic children [@jabere2018xperiment]. Even the effect of age is unstable in this sample of studies. Whereas it is quite clear that children's NWR scores increase in the Italian data, age effects are less stable among Tsimane' children. Therefore, all analyses in the present study are descriptive and should be considered exploratory.

```{r prevlit-fig, echo = FALSE, ig.pos = "T!", fig.width=8, fig.height=5.5, fig.cap = "NWR scores as a function of age (in years) and item length for comparable studies. Arabic data  from Jaber et al. (2018); Italian data from Piazzalunga et al. (2019); Tsimane' from Cristia et al. (2020); Yélî Dnye data from the present study."}

@@

A second list of candidate items was generated in 2019 by selecting complex consonants and systematically crossing them with all the vowels in the Yélî Dnye inventory to produce consonant-vowel monosyllabic forms. As before, items were automatically excluded if they appeared in the dictionary. Additionally, since perceiving vowel length in isolated monosyllables is challenging, any item that had a short/long lexical neighbor was excluded. Because there is still much to discover about the phonology and phonetics of Yélî Dnye [@levinsonYDgrammar], it was also possible that we initially generated items with illegal, but currently undocumented constraints. Therefore, we made sure that the precise consonant-vowel sequence occurred in some real word in the dictionary (i.e., that there was a longer word included the monosyllable as a subsequence). These candidates were then presented to one informant, for a final check that they did not mean anything. Together with the 2018 selection, they were recorded, based on their orthographic forms, using a Shure SM10A XLR dynamic headband microphone and an Olympus WS-832 stereo audio recorder (using an XLR to mini-jack adapter) by the same informant, monitored by the second author for clear production of the phonological target. The complete recorded list was finally presented to two more informants, who were able to repeat all the items and who confirmed there were no real words present. Despite these checks, one monosyllable was ultimately frequently identified as a real word in the resulting data (intended "yî" /yɯ/; identified as "yi" /yi/, *tree*). This item is removed from the analyses below.

@@

kable(stims.wide,
  "latex", booktabs = T,
  caption="NWR stimuli in orthographic and phonological representations.",
  row.names =F) %>%
  kable_styling() %>%
  add_header_above(c("Practice" = 2, "Monosyll" = 2,
    "Bisyll" = 2, "Trisyll" = 2, "Tetrasyll" = 2))

@@

A Praat script [@Praat] was written to randomize this list 20 times, and split it into two sublists, to generate 40 different elicitation sets. The 40 elicitation sets are available online from [osf.io/5qspb](https://osf.io/5qspb). The split had the following constraints:

@@

@@

Finally, we looked into the frequency with which mispronunciations resulted in real words. In fact, two thirds of incorrect repetitions were recognizable as real words or phrases in Yélî Dnye or English: `r round(sum(!is.na(trnsc.inc$meaning) & trnsc.inc$meaning != "0")/sum(!is.na(trnsc.inc$mispronunciation))*100)`%. This type of analysis is seldom reported---we could only find one comparison point: @castro1998 found that illiterate Portuguese adults' NWR mispronunciations resulted in real words in `r round(53/475*100,2)`% of cases, whereas literate participants did so in only `r round(2/117*100,2)`% of cases. The percentage we observe here is much higher than reported in Castro and colleagues' study, but we do not know whether age, language, or even test structure explains this difference.

>>>>

abstract: >
   In nonword repetition (NWR) studies, participants are presented auditorily with an item that is phonologically legal but lexically meaningless in their language, and asked to repeat this item as closely as possible. NWR scores are thought to reflect some aspects of phonological development, saliently a perception-production loop supporting flexible production patterns. In this study, we report on NWR results among children learning Yélî Dnye, an isolate spoken on Rossel Island in Papua New Guinea. This study contributes to four lines of research. First, we document that non-word items containing typologically frequent sounds are repeated without changes more often that non-words containing typologically rare sounds, above and beyond any within-language frequency effects. Second, we document rather weak effects of item length, contributing to mounting research suggesting that length effects may be language- or population-specific. Third, we do not find strong individual variation effects in this population. Fourth, we discuss how NWR scores can be compared across studies, languages, and populations, in order to shed light on the factors universally structuring variation  in language development.  Together, these data provide a unique view of online phonological processing in an understudied language while making preliminary connections between language development and cross-linguistic features.

@@

Children's perception and production of phonetic and phonological units continues developing well beyond the first year of life, even extending into middle childhood [e.g., @hazan2000development]. Much of the evidence for later phonological development comes from nonword repetition (NWR) tasks. In a NWR task, participants hear a short word-like form that is phonologically legal but lexically meaningless in the language(s) they are learning. After hearing this non-word, the participant's task is to try to immediately and precisely repeat it. NWR scores are thought to reflect long-term phonological knowledge (to perceive the item precisely despite not having heard it before) as well as online phonological working memory (to encode the item in the interval between hearing it and saying it back) and flexible production patterns (to produce the item precisely despite not having pronounced it before).

NWR has been used to seek answers to a variety of theoretical questions, including what the links are between phonology, working memory, and the lexicon [@bowey2001nonword], and how extensively phonological constraints found in the lexicon affect online production [@gallagher2014identity]. NWR is also frequently used in applied contexts, notably as a diagnostic tool for language delays and disorders [@estes2007differences]. Since non-words can be generated in any language, it has attracted the attention of researchers working in multilingual and linguistically diverse environments, particularly in Europe [@meir2016disentangling;@is08042009language].

<!-- a2m No, I didn't propose that we would use NWR because I wanted to contribute to our understanding of the NWR task in new settings, but rather because it's quite a standard task so it could be useful for us documenting phonological development in Rossel. To make this clearer, I moved it down and integrated it with the four contributions below -->

In the present study, we use NWR to investigate the phonological development of children learning Yélî Dnye, an isolate language spoken in Papua New Guinea (PNG) that has a large and unusually dense phonological inventory. The study was designed to contribute to four aspects of our understanding of phonological development.

First, we included a subset of non-word items with typologically rare and/or challenging sounds to ask whether these rare sounds are disadvantaged in the perception-production loop involved in NWR. Previous work using NWR has preferred relatively universal and early-acquired phonemes [with the possible exception of @gallagher2014identity], likely as a way to separate phoneme pronunciation from broader syllable structure and word-level prosodic effects [@gallon2007non]. Here, we investigate repetition of non-word items containing cross-linguistically common and cross-linguistically rare phonetic targets.

Second, we varied the length (in syllables) of non-words to contribute to growing research looking at the impact of word length on NWR repetition, and what this may reflect about phonological development. Our reading of previous NWR research is that there are variable effects of length between populations. For instance, @jabere2018xperiment reports an average of 95% correct repetition for items 1 or 2 syllables long among Arabic-speaking children at about 5 years of age, but 55% for items 3 or 4 syllables long. In contrast, @piazzalunga2019articulatory observe nearly no change in performance in similarly-aged Italian learners, with a score of 84% for 1-2 syllables versus 85% for 3-4 syllables. It is possible that differences are due to a host of variables, including the modal length of words in the language and/or in child-directed speech in that culture. That said, there could be other causal pathways: Research on adults suggests that illiterate Portuguese speakers repeat monosyllabic non-words just as accurately as literate adults, whereas scores are much lower among illiterate than literate speakers for items 3 or 4 syllables long [@de2004degree]. Given that both groups of adults speak the same language (Brazilian Portuguese), then perhaps differences in repetition accuracy reveal differences in how flexible the perception-production loop is. Given the paucity of evidence looking at this question, we do not approach this issue within a hypothesis-testing framework but sought instead to provide one more piece of data on the question, which may be re-used in future meta- or mega-analytic approaches.

Third, there are ongoing discussions as to what the key factors structuring individual variation are. Although the ideal systematic review is missing, a recent paper comes close with a rather extensive review on the literature looking at correlations between NWR scores and a variety of child-level variables [@farabolini2021nonword]. In a nutshell, most evidence is mixed, suggesting that consistent individual variation effects may be small, and more data is needed to estimate their true size. For this reason, we descriptively report association strength between NWR scores and child age, sex, birth order, and maternal education. Based on previous work, we looked at potential changes with age [@farmani2018normalization; @kalnak2014nonword; @vance2005speech]<!--, and differences relating to bilingual language exposure [@brandeker2015language; @meir2016disentangling; @meir2017independent]-->. Previous work typically finds no significant differences as a function of maternal education [e.g., @farmani2018normalization; @kalnak2014nonword; @meir2017independent; @balladares2016socio] or child gender [@chiat2007preschool]. Although past research has not often investigated potential effects of birth order on NWR, there is a sizable literature on these effects in other language tasks [@havron2019effect], and therefore we report on these too.

<!-- a2m do you remember: why did we remove bilingual exposure? -->

Fourth, these data contribute to the small literature using this task with non-Western, non-urban populations, speaking a language with a moderate to large phonological inventory [see @maddieson2005correlating for a broad classification of languages based on inventory size]. Indeed, NWR has seldom been used outside of Europe and North America [with exceptions including @gallagher2014identity; @cristia2020infant], outside urban settings [except for in @cristia2020infant], nor with languages having large phonological inventories. There are no theoretical reasons to presume that the technique will not generalize to these new conditions. That said, @cristia2020infant recently reported relatively lower NWR scores among the Tsimane', a non-Western rural population, interpreting these findings as consistent with the hypothesis that lower levels of infant-directed speech and/or low prevalence of literacy in a population could lead to population-level differences in NWR scores. In view of these results, it is important to bear in mind that NWR is a task developed in countries where literacy is widespread, and it is considered an excellent predictor of reading, for instance better than rhyme awareness [e.g., @gathercole1991differentiating]. Therefore, it may not be a general index of phonological development, but more specifically reflect certain skills. Indeed, @cristia2020infant present the task as being a good index of the development of "short-hand-like" representations specifically, which could thus miss, for example, more holistic representations.

Before going into the details of our study design we first give an overview of Yélî Dnye phonology as well as a brief ethnographic review of the developmental environment on Rossel Island. As discussed above, NWR has been almost exclusively used in urban, industrialized populations, so we provide this additional ethographic information to contextualize the adaptations we have made in running the task and in gathering age and other demographic information, compared to what is typical in urban, industrialized settings.

@@

In total, Yélî Dnye uses 90 distinctive segments (not including an additional 3 rarely used consonants), far outstripping the phonemic inventory size of other documented Papuan languages [@foley1986papuan; @levinsonYDgrammar] and includes at least two contrasts that had not previously been documented elsewhere [labial-coronal double-articulations with dental vs. post-alveolar coronal placement for both oral and nasal stops, @ladefoged1996sounds; @maddiesonIPphoneticsYD].

Most words in Yélî Dnye are bisyllabic (~50%), with monosyllabic words (~40%) appearing most commonly after that, and with tri-and-above syllabic words appearing least frequently [~10%; based on >5800 lexemes in the most recent Yélî Dnye dictionary at the time of writing; @levinsonYDgrammar]. The vast majority of syllables use a CV format. A small portion of the lexicon features words with a final closed syllable, limited to codas of -/m/, -/p/, or -/j/ (e.g., "ndap" (Spondylus shell) /ṇṭæp/). However, in spontaneous speech an epenthetic /ɯ/ is often inserted after word-final coda, yielding a CV.CV structure instead (e.g., "ndap" pronounced "ndapî" /'ṇṭæ.pɯ/). This process is used frequently with English loan words that have a coda (e.g., "ponî" (phone) /'pɔ.nɯ/). An even smaller portion of the lexicon features words starting with a vowel (e.g., "ala" (here) /æ.'læ/), but these are limited to /æ/-. Finally, the lexicon features a handful of single-vowel grammatical morphemes (see @levinsonYDgrammar for details).

@@

Most speakers of Yélî Dnye grow up speaking it monolingually until they begin attending school around the age of 7 years; school instruction is in English. While monolingual Yélî Dnye upbringing is common, multilingual families are not unusual, particularly in the region around the Catholic Mission---the same region in which the current data were collected---where there is a higher incidence of married-in mothers from other islands [@brownIPchildrearing]. Children in these multilingual families grow up speaking Yélî Dnye plus English, Tok Pisin, and/or other language(s) from the region.

<!-- a2m I thought in the first couple of years they combined English and Yélî at school? -->

@@

Laying 250 nautical miles off the coast of mainland PNG and surrounded by a barrier reef, transport to and from the island is both infrequent and irregular. There is no electricity system (visiting researchers use solar panels) or motor vehicles on the island, save dinghys with outboard motors that are primarily reserved for medical transport when there is diesel available. There is a medical radio connection at the health clinic, located at the Catholic Mission, via which news and messages are often passed to and from mainland PNG. There is also a cellular tower on the island, though it is often inoperative, usually for multiple months or years at a time. As such, access to news and outside connections, including international phone calls and digital exchanges that require significant data transfer, are typically unavailable. Our data collection is therefore limited to the duration of the researchers' on-island visits.

Despite restricted outside contact, formal education is a priority for many Yélî families. A recent study surveying more than 40 parents of young children in this region of the island found that nearly all parents completed 6 or more years of education, with 65% of mothers and 49% of fathers completing the full range of locally available education (~9 years), with around half of those parents having pursued an additional 2+ years of education on other islands in the region or on mainland PNG [@casillas2020early]. While there are a handful of local schools around the island, these may be well out of walking distance for many children (i.e., more than 1 hour on foot or by canoe each day). It is very common for households that are situated close to a school to billet their school-aged relatives during the weekdays (sometimes the weekends too, if the child's home is very far) for long segments of each school year. Combined with practices of collective child guardianship within each close-knit hamlet, the practical consequence of this billeting for researchers is that adult consent can often come from a combination of aunts, uncles, adult cousins, and grandparents standing in for the child's biological parents. On top of this, child assent is culturally pertinent, as child independence is encouraged and respected from toddlerhood onward [@brownIPchildrearing].^[For these reasons, we were unable to record precise details for a few children, including child date of birth, parent age in years, and parent education.] Finally, as mentioned, the language of schooling in this region is English, and so the majority of teens and adults on the island speak some English---many speak it very fluently.

<!-- a2m why do we explain this: Finally, as mentioned, the language of schooling in this region is English, and so the majority of teens and adults on the island speak some English---many speak it very fluently. -->

While work on Yélî language development is growing [e.g., @brown2011cultural; @brown2014interactional; @brownIPchildrearing; @casillas2020early; @liszkowski2012prelinguistic], our knowledge of children's linguistic development is quite limited, and research into their phonological development in particular has only just begun [e.g., @peuteIPconsonants]. What we do know about the early language environment is that, while direct speech to children under 3;0 is relatively infrequent throughout the day [@casillas2020early], shared caregiving practices and a near-universal fondness for social interaction with young children ensure that speech directed to children comes from all types of speakers: women, men, and other children [@bunceURcrosscultural; @casillas2020early]. While speech from women predominates in young children's language environments on Rossel, as it does elsewhere [@bergelsoncasillas2019what; @bunceURcrosscultural], there is a significant and marked increase in child-directed speech from other children as infants get older [@bunceURcrosscultural; @casillas2020early]. The latter increase is attributed to the fact that, starting around age 2, children often spend large swaths of the day playing, swimming, and foraging for fruit, nuts, and shellfish in large (~10 members) independent and mixed-age child play groups [@brownIPchildrearing; @casillas2020early].

<!-- a2m including Bunce here may be ambiguous, and it could be read to indicate that speech from other children increases in other cultures as well: there is a significant and marked increase in child-directed speech from other children as infants get older [@bunceURcrosscultural; @casillas2020early] -->

@@

In a basic NWR task, the participant listens to a production of a word-like form, such as /bilik/, and then repeats back what they heard without changing any phonological feature that is contrastive in the language. For instance, in English, a response of [bilig] or [pilik] would be scored as incorrect; a response [biːlik], where the vowel is lengthened without change of quality would be scored as correct, because English does not have a general feature of contrastive vowel length. There is some variation in how past NWR studies have designed the presentation procedure and structure of items. For example, while items are often presented orally by the experimenter [@torrington2015non], an increasing number of studies have turned instead to playing back pre-recorded stimuli in order to increase control in stimulus presentation [@brandeker2015language]. Additionally, while some studies have used 10-15 non-words [e.g., @cristia2020infant], others have employed up to 46 unique items [@piazzalunga2019articulatory]. Authors also often modulate structural complexity, typically measured in terms of item length (measured in number of syllables) and/or syllable structure [open as opposed to closed syllables, @gallon2007non].

Previous work typically steers clear of articulatorily and/or acoustically challenging sounds, but we included some in our experiment to more adequately represent Yélî Dnye's phonology and to contribute data on whether this affects repetition. We ultimately used a relatively large number of items that would enable us to explore both variation in structural complexity and in more vs. less challenging sounds. However, aware that this large item inventory might render the task longer and more tiresome, we split items across children (see below). Naturally, designing the task in this way may make the study of individual variation within the population more difficult because different children are exposed to different items. However, as discussed above, effects of individual differences in NWR are probably relatively small, and thus we reasoned that they would not be detectable with the sample size that we could collect during our short visit. That said, we contribute to the literature by also reporting descriptive analyses of individual variation that could potentially be integrated in meta- or mega-analytic efforts.

@@

- Does the cross-linguistic frequency of sounds in the stimuli predict NWR scores? Are rarer sounds more often substituted by commoner sounds?
- How do NWR scores change as a function of item length in number of syllables?
- Is individual variation in NWR scores attributable to child age, sex, birth order, and/or maternal education?

We had considered boosting the interpretational value of this evidence by announcing our analysis plans prior to conducting them. However, we realized that even pre-registering an analysis would be equivocal because we would not have enough power to look at all relationships of interest, in many cases possibly not enough to detect any of the known effects, given their variability across studies. To illustrate this issue, we portray in Figure 1 studies in which children's NWR scores were gathered between 4 and 12 years of age, and reported separately for items that are relatively short (1-2 syllables) versus longer items (3-4 syllables). As discussed above,  the effect of stimulus length is minuscule among Italian children [@piazzalunga2019articulatory], but considerable among Arabic children [@jabere2018xperiment]. Even the effect of age is unstable in this sample of studies. Whereas it is quite clear that children's NWR scores increase in the Italian data [@piazzalunga2019articulatory], age effects are less stable among Tsimane' children [@cristia2020infant]. Therefore, all analyses in the present study are descriptive and should be considered exploratory.

```{r prevlit-fig, echo = FALSE, ig.pos = "T!", fig.width=8, fig.height=5.5, fig.cap = "NWR scores as a function of age (in years) and item length for comparable studies. Arabic data  from Jaber-Awida (2018); Italian data from Piazzalunga et al. (2019); Tsimane' from Cristia et al. (2020); Yélî Dnye data from the present study."}

@@

A second list of candidate items was generated in a second trip to the island in 2019, when data were collected, by selecting complex consonants and systematically crossing them with all the vowels in the Yélî Dnye inventory to produce consonant-vowel monosyllabic forms. As before, items were automatically excluded if they appeared in the dictionary. Additionally, since perceiving vowel length in isolated monosyllables is challenging, any item that had a short/long lexical neighbor was excluded. Because there is still much to discover about the phonology and phonetics of Yélî Dnye [@levinsonYDgrammar], it was also possible that we initially generated items with illegal, but currently undocumented constraints. Therefore, we made sure that the precise consonant-vowel sequence occurred in some real word in the dictionary (i.e., that there was a longer word included the monosyllable as a subsequence). These candidates were then presented to one informant, for a final check that they did not mean anything. Together with the 2018 selection, they were recorded, based on their orthographic forms, using a Shure SM10A XLR dynamic headband microphone and an Olympus WS-832 stereo audio recorder (using an XLR to mini-jack adapter) by the same informant, monitored by the second author for clear production of the phonological target. The complete recorded list was finally presented to two more informants, who were able to repeat all the items and who confirmed there were no real words present. Despite these checks, one monosyllable was ultimately frequently identified as a real word in the resulting data (intended "yî" /yɯ/; identified as "yi" /yi/, *tree*). This item is removed from the analyses below.

@@

kable(stims.wide,
  "latex", booktabs = T,
  caption="NWR stimuli in orthographic (Orth.) and phonological (Phon.) representations.",
  row.names =F) %>%
  kable_styling() %>%
  add_header_above(c("Practice" = 2, "Monosyll" = 2,
    "Bisyll" = 2, "Trisyll" = 2, "Tetrasyll" = 2))

@@

A Praat script [@Praat] was written to randomize this list 20 times, and split it into two sublists, to generate 40 different elicitation sets. The 40 elicitation sets are available online from [osf.io/dtxue/](https://osf.io/dtxue/). The split had the following constraints:

@@

# read.csv("nwr-demo.csv",header=T)->demo
# demo=demo[,-c(20,21)] #remove DOBs 
# write.csv(demo,"nwr-demo.csv")

@@

Finally, we looked into the frequency with which mispronunciations resulted in real words. In fact, two thirds of incorrect repetitions were recognizable as real words or phrases in Yélî Dnye or English: `r round(sum(!is.na(trnsc.inc$meaning) & trnsc.inc$meaning != "0")/sum(!is.na(trnsc.inc$mispronunciation))*100)`%. This type of analysis is seldom reported---we could only find one comparison point: @castro1998illiterate found that illiterate European Portuguese adults' NWR mispronunciations resulted in real words in `r round(53/475*100,2)`% of cases, whereas literate participants did so in only `r round(2/117*100,2)`% of cases. The percentage we observe here is much higher than reported in Castro and colleagues' study, but we do not know whether age, language, or even test structure explains this difference.
